Water desalination using polyelectrolyte hydrogel. Gibbs ensemble modelling

### 2022-07-06 gibbs.tex
presubmission version

\paragraph{Compression in open system.}
\begin{figure*}[t]
\subfloat[The gel partial pressure vs gel molar volume \label{fig: PV}]
{\includegraphics[width=0.48\textwidth]{figures/fig_PV_pub}}
\hspace{0.02\textwidth}
\subfloat[Supernate salinity vs gel molar volume\label{fig: CV}]
{\includegraphics[width=0.48\textwidth]{figures/fig_CV_pub}}
\caption{
The compression of the gel in open system (solid lines) and in closed system (dotted lines). 
Each solid curve corresponds to different salinity of the reservoir $\cs$ (see legend). Shadowed area limit the states with applied pressure below zero and above 5 bar.\label{fig: PV and CV}}
\todoi{replace $V$ by $\Vgel$ in Figures}
\end{figure*}
First, we run a set of simulations modeling the gel compression in \emph{open system}, \ie in equilibrium with a big bath of certain salinity, $\cs$. 
The simulations are run for a set of different simulation box volumes. 
Each simulation returns the averaged pressure, $P$, and the number of $\cl$ ions, which are present in the simulation box, $\ncl$.

In order to get the partial pressure of the gel, we substitute from $P$ the osmotic pressure of ions in the reservoir: $\Pgel=P - \Pout$.
$\Pout$, in turn, we calculate by running a separate simulation of a reservoir, containing ionic gas in equilibrium with the bath of the same salinity $\cs$. 
The gel partial pressure, $\Pgel$, is the pressure that needs to be applied to the gel via a solvent permeable filter to compress the gel to a molar-specific volume, $\Vgel$.

The dependencies of $\Pgel$ on the gel molar volume $\Vgel$ are present in Figure \ref{fig: PV} for a set of different salinities. \todo{consider treplacing $\Pgel$ by $\Pi$}
These dependencies (the case of \emph{open system}) are represented by solid lines. 
For example, the blue solid line illustrates the compression (or swelling) of the gel in equilibrium with a reservoir of salinity, $\cs=0.063$ mol/l. 
The points where the pressure equals zero, $\Pgel=0$, (indicated by filled circles) are the gel \emph{free swelling equilibrium} states. 
These states shift towards smaller volumes with increase of salinity. 
In general, the increase of salinity shift all the $\Pgel(V)$ curves towards smaller volumes.
This effect is well known and is typical for all branched \emph{strong} polyelectrolytes. 
It is caused by the decrease of ions osmotic pressure and by a screening of electrostatic interactions \cite{Zhulina2000, Landsgesel2020a}.
\footnote{The salinity dependence of the size of \emph{weak} polyelectrolyte gel is in general non-monotonic. We discuss this case in \cite{Rud2018}.}

\begin{figure*}[t]
\subfloat[Pressure vs gel molar volume \label{fig: NV}]
{\includegraphics[width=0.49\textwidth]{figures/fig_NV_pub}}
\subfloat[Salinity vs gel molar volume\label{fig: CN}]
{\includegraphics[width=0.495\textwidth]{figures/fig_NVbox_mu_pub}}
\caption{The compression of the gel in open system (solid lines) and in closed
system (dotted lines). Each solid curve corresponds to different salinity
of the reservoir $\cs$ (see legend). Shadowed area limit the states
with applied pressure below zero and above 5 bar.\label{fig: NV and CN}}
%\todoi{Replace y-axis labels by $\ncl$}
\end{figure*}

@@

\todo{we could get rid of saying 'gel molar volume' if we state elsewhere that all the values are calculated per one mol of the gel}

@@

Such estimations are provided in the fifth column of the table, for five triplets of values $\cs^f$, $\cs^p$, and $\cs^b$.
$W_{\mathrm{sim}}$ is the energy calculated by integration of $\Pgel(\Vgel)$ dependencies~--- sum of corresponding comressions in closed and open systems, $W_{\mathrm{id}}$~--- the estimate based on the equation \refeq{eq:SEC} and $R_w$~--- is the corresponding recovery ratio.
Note that calculating the $W_{\mathrm{sim}}$ we accounted for only the work done on the gel during compression, whereas the work done by the gel itself during swelling was not taken into account.

\section{Conclusions}
\begin{enumerate}
\item We have modeled the compression of the gel in thermodynamic equilibrium
with the reservoir of limited amount of water.
\item We have shown that in case of gel made of strong polyelectrolyte,
the compression always lead to decrease of the surrounding salinity.
\item We have modeled the process of water desalination combining two processes:
(1) the swelling of the gel in open system, exchanging ions with big
reservoir at constant salinity. (2) the compression of the gel in
closed system, when the gel affect the salinity in surrounding solution.
\item We estimated the energy needed to produce one liter of water and have
show that the method at least theoretically may compete with the modern
technologies used in industry nowadays.
\end{enumerate}

@@

\newpage
\listoftodos

@@

\begin{tabular*}{1\textwidth}{@{\extracolsep{\fill}}ll|lc|c|l|l|ll}
\multicolumn{1}{l|}{$c_{s}^{0}$, mM} & $c_{s}^{5}$, mM & \multicolumn{1}{l|}{$n_{\cl}^{0}$, mM} & $n_{\cl}^{5}$, mM & $\Delta v$, l & $R_{w}$ & $|W$\textbar , J/L & \multicolumn{2}{c}{$W^{id}$, J/L}\tabularnewline
\multirow{2}{*}{{\small{}91.466}} & \multirow{2}{*}{} & \multicolumn{2}{l|}{{\small{}$57.083\pm0.1215\quad\longrightarrow$}} & \multirow{2}{*}{{\small{}2.74}} & \multirow{2}{*}{} & \multirow{2}{*}{{\small{}$95.4\pm1.86$}} &  & \multirow{2}{*}{}\tabularnewline
 &  & \multicolumn{2}{r|}{{\small{}$71.751\pm0.024$}} &  &  &  & \multirow{8}{*}{{\small{}\hspace{-1em}}\textcolor{teal}{\small{}$\left.\begin{array}{l}
\end{array}\right\rbrace $$\rotatebox{90}{\hspace{-3.8em}\ensuremath{\begin{array}{cc}
W^{id}= & (52.9)\\
\cline{1-7} \cline{2-7} \cline{3-7} \cline{4-7} \cline{5-7} \cline{6-7} \cline{7-7} 
\multicolumn{2}{l|}{{\small{}$89.414\pm0.2286\quad\longrightarrow$}} & \multirow{2}{*}{{\small{}56.898}} & \multirow{2}{*}{} & \multirow{2}{*}{{\small{}2.72}} & \multirow{2}{*}{} & \multirow{2}{*}{{\small{}$109.1\pm1.72$}} &  & \multirow{2}{*}{}\tabularnewline
\multicolumn{2}{r|}{{\small{}$73.633\pm0.029$}} &  &  &  &  &  &  & \tabularnewline
\cline{1-7} \cline{2-7} \cline{3-7} \cline{4-7} \cline{5-7} \cline{6-7} \cline{7-7} 
\multirow{2}{*}{\textcolor{teal}{\small{}63.93}} & \multirow{2}{*}{} & \multicolumn{2}{l|}{\textcolor{teal}{\small{}$37.281\pm0.083\quad\longrightarrow$}} & \multirow{2}{*}{\textcolor{teal}{\small{}3.26}} & \multirow{2}{*}{\textcolor{teal}{\small{}0.54}} & \multirow{2}{*}{\textcolor{teal}{\small{}$100.9\pm1.69$}} &  & \tabularnewline
 &  & \multicolumn{2}{r|}{\textcolor{teal}{\small{}$50.579\pm0.0134$}} &  &  &  &  & \multirow{8}{*}{{\small{}\hspace{-1em}}\textcolor{olive}{\small{}$\left.\begin{array}{l}
\end{array}\right\rbrace $$\rotatebox{90}{\hspace{-3.8em}\ensuremath{\begin{array}{cc}
\cline{1-7} \cline{2-7} \cline{3-7} \cline{4-7} \cline{5-7} \cline{6-7} \cline{7-7} 
\multicolumn{2}{l|}{\textcolor{teal}{\small{}$62.052\pm0.148\quad\longrightarrow$}} & \multirow{2}{*}{\textcolor{teal}{\small{}37.172}} & \multirow{2}{*}{} & \multirow{2}{*}{\textcolor{teal}{\small{}3.18}} & \multirow{2}{*}{} & \multirow{2}{*}{\textcolor{teal}{\small{}$107.4\pm1.37$}} &  & \tabularnewline
\multicolumn{2}{r|}{\textcolor{teal}{\small{}$48.206\pm0.0157$}} &  &  &  &  &  &  & \tabularnewline
\cline{1-7} \cline{2-7} \cline{3-7} \cline{4-7} \cline{5-7} \cline{6-7} \cline{7-7} 
\multirow{2}{*}{\textcolor{olive}{\small{}44.915}} & \multirow{2}{*}{} & \multicolumn{2}{l|}{\textcolor{olive}{\small{}$23.747\pm0.0572\quad\longrightarrow$}} & \multirow{2}{*}{\textcolor{olive}{\small{}3.82}} & \multirow{2}{*}{\textcolor{olive}{\small{}0.53}} & \multirow{2}{*}{\textcolor{olive}{\small{}$106.7\pm1.51$}} &  & \tabularnewline
 &  & \multicolumn{2}{r|}{\textcolor{olive}{\small{}$35.911\pm0.0101$}} &  &  &  & \multirow{8}{*}{{\small{}\hspace{-1em}}\textcolor{red}{\small{}$\left.\begin{array}{l}
\end{array}\right\rbrace \rotatebox{90}{\hspace{-3.8em}\ensuremath{\begin{array}{cc}
W^{id}= & 43.8\ {\color{red}{\scriptstyle (41.0)}}\\
\cline{1-7} \cline{2-7} \cline{3-7} \cline{4-7} \cline{5-7} \cline{6-7} \cline{7-7} 
\multicolumn{2}{l|}{\textcolor{olive}{\small{}$43.463\pm0.116\quad\longrightarrow$}} & \multirow{2}{*}{\textcolor{olive}{\small{}24.272}} & \multirow{2}{*}{} & \multirow{2}{*}{\textcolor{olive}{\small{}3.91}} & \multirow{2}{*}{} & \multirow{2}{*}{\textcolor{olive}{\small{}$106.4\pm1.13$}} &  & \tabularnewline
\multicolumn{2}{r|}{\textcolor{olive}{\small{}$30.795\pm0.0078$}} &  &  &  &  &  &  & \tabularnewline
\cline{1-7} \cline{2-7} \cline{3-7} \cline{4-7} \cline{5-7} \cline{6-7} \cline{7-7} 
\multirow{2}{*}{\textcolor{red}{\small{}31.928}} & \multirow{2}{*}{} & \multicolumn{2}{l|}{\textcolor{red}{\small{}$14.658\pm0.0368\quad\longrightarrow$}} & \multirow{2}{*}{\textcolor{red}{\small{}4.20}} & \multirow{2}{*}{\textcolor{red}{\small{}0.52}} & \multirow{2}{*}{\textcolor{red}{\small{}$107.9\pm1.42$}} &  & \tabularnewline
 &  & \multicolumn{2}{r|}{\textcolor{red}{\small{}$25.297\pm0.0061$}} &  &  &  &  & \multirow{8}{*}{{\small{}\hspace{-1em}}\textcolor{orange}{\small{}$\left.\begin{array}{l}
\end{array}\right\rbrace \rotatebox{90}{\hspace{-3.8em}\ensuremath{\begin{array}{cc}
\cline{1-7} \cline{2-7} \cline{3-7} \cline{4-7} \cline{5-7} \cline{6-7} \cline{7-7} 
\multicolumn{2}{l|}{\textcolor{red}{\small{}$30.029\pm0.0839\quad\longrightarrow$}} & \multirow{2}{*}{\textcolor{red}{\small{}14.549}} & \multirow{2}{*}{} & \multirow{2}{*}{\textcolor{red}{\small{}$\begin{array}{c}
\end{array}$}} & \multirow{2}{*}{} & \multirow{2}{*}{\textcolor{red}{\small{}$\begin{array}{c}
18.585\pm0.0036\,\\
\end{array}$}} &  &  &  &  &  &  & \tabularnewline
\cline{1-7} \cline{2-7} \cline{3-7} \cline{4-7} \cline{5-7} \cline{6-7} \cline{7-7} 
\multirow{2}{*}{\textcolor{orange}{\small{}22.301}} & \multirow{2}{*}{} & \multicolumn{2}{l|}{\textcolor{orange}{\small{}$8.838\pm0.0262\quad\longrightarrow$}} & \multirow{2}{*}{\textcolor{orange}{\small{}$\begin{array}{c}
\end{array}$}} & \multirow{2}{*}{\textcolor{orange}{\small{}0.53}} & \multirow{2}{*}{\textcolor{orange}{\small{}$\begin{array}{c}
\text{17.945}\pm\text{0.0041}\\
\end{array}$}} &  &  &  & \multirow{7}{*}{{\small{}\hspace{-1em}}\textcolor{magenta}{\small{}$\left.\begin{array}{l}
\end{array}\right\rbrace \rotatebox{90}{\hspace{-3em}\ensuremath{\begin{array}{cc}
\cline{1-7} \cline{2-7} \cline{3-7} \cline{4-7} \cline{5-7} \cline{6-7} \cline{7-7} 
\multicolumn{2}{l|}{\textcolor{orange}{\small{}$20.743\pm0.0638\quad\longrightarrow$}} & \multirow{2}{*}{\textcolor{orange}{\small{}8.827}} & \multirow{2}{*}{} & \multirow{2}{*}{\textcolor{orange}{\small{}$4.71$}} & \multirow{2}{*}{} & \multirow{2}{*}{\textcolor{orange}{\small{}$110.8\pm0.77$}} &  & \tabularnewline
\multicolumn{2}{r|}{\textcolor{orange}{\small{}$11.082\pm0.002$}} &  &  &  &  &  &  & \tabularnewline
\cline{1-7} \cline{2-7} \cline{3-7} \cline{4-7} \cline{5-7} \cline{6-7} \cline{7-7} 
\multirow{2}{*}{\textcolor{magenta}{\small{}10.904}} & \multirow{2}{*}{} & \multicolumn{2}{l|}{\textcolor{magenta}{\small{}$3.002\pm0.0098\quad\longrightarrow$}} & \multirow{2}{*}{\textcolor{magenta}{\small{}6.08}} & \multirow{2}{*}{\textcolor{magenta}{\small{}0.55}} & \multirow{2}{*}{\textcolor{magenta}{\small{}$106.9\pm1.05$}} &  & \tabularnewline
 &  & \multicolumn{2}{r|}{\textcolor{magenta}{\small{}9.011$\pm$0.0019}} &  &  &  &  & \tabularnewline
\cline{1-7} \cline{2-7} \cline{3-7} \cline{4-7} \cline{5-7} \cline{6-7} \cline{7-7} 
\multicolumn{2}{l|}{\textcolor{magenta}{\small{}$9.834\pm0.0461\quad\longrightarrow$}} & \multirow{2}{*}{\textcolor{magenta}{\small{}3.119}} & \multirow{2}{*}{} & \multirow{2}{*}{\textcolor{magenta}{\small{}5.78}} & \multirow{2}{*}{} & \multirow{2}{*}{\textcolor{magenta}{\small{}$119.4\pm0.97$}} &  & \tabularnewline
\multicolumn{2}{r|}{\textcolor{magenta}{\small{}$3.862\pm0.0014$}} &  &  &  &  &  &  & \tabularnewline

\caption{ORIGINAL TABLE (to be removed)
}
\end{table}

>>>>

\paragraph{Compression in open system.}
\begin{figure*}[t]
\subfloat[The gel partial pressure vs gel molar volume \label{fig: PV}]
{\includegraphics[width=0.48\textwidth]{figures/fig_PV_pub}}
\hspace{0.02\textwidth}
\subfloat[Supernate salinity vs gel molar volume\label{fig: CV}]
{\includegraphics[width=0.48\textwidth]{figures/fig_CV_pub}}
\caption{
The compression of the gel in open system (solid lines) and in closed system (dotted lines). 
Each solid curve corresponds to different salinity of the reservoir $\cs$ (see legend). Shadowed area limit the states with applied pressure below zero and above 5 bar.\label{fig: PV and CV}}
\end{figure*}
First, we run a set of simulations modeling the gel compression in \emph{open system}, \ie in equilibrium with a big bath of certain salinity, $\cs$. 
The simulations are run for a set of different simulation box volumes. 
Each simulation returns the averaged pressure, $P$, and the number of $\cl$ ions, which are present in the simulation box, $\ncl$.

In order to get the partial pressure of the gel, we substitute from $P$ the osmotic pressure of ions in the reservoir: $\Pgel=P - \Pout$.
$\Pout$, in turn, we calculate by running a separate simulation of a reservoir, containing ionic gas in equilibrium with the bath of the same salinity $\cs$. 
The gel partial pressure, $\Pgel$, is the pressure that needs to be applied to the gel via a solvent permeable filter to compress the gel to a specific molar volume, $\Vgel$. 
By gel molar volume $\Vgel$ we assume the volume of the gel calculated per on mole of gel monomer units.

The dependencies of $\Pgel$ on $\Vgel$ are present in Figure \ref{fig: PV} for a set of different salinities. 
These dependencies (the case of \emph{open system}) are represented by solid lines. 
For example, the blue solid line illustrates the compression (or swelling) of the gel in equilibrium with a reservoir of salinity, $\cs=0.063$ mol/l. 
The points where the pressure equals zero, $\Pgel=0$, (indicated by filled circles) are the gel \emph{free swelling equilibrium} states. 
These states shift towards smaller volumes with increase of salinity. 
In general, the increase of salinity shift all the $\Pgel(V)$ curves towards smaller volumes.
This effect is well known and is typical for all branched \emph{strong} polyelectrolytes. 
It is caused by the decrease of ions osmotic pressure and by a screening of electrostatic interactions \cite{Zhulina2000, Landsgesel2020a}.
\footnote{The salinity dependence of the size of \emph{weak} polyelectrolyte gel is in general non-monotonic. We discuss this case in \cite{Rud2018}.}

\begin{figure*}[t]
\subfloat[Pressure vs gel molar volume \label{fig: NV}]
{\includegraphics[width=0.49\textwidth]{figures/fig_NV_pub}}
\subfloat[Salinity vs gel molar volume\label{fig: CN}]
{\includegraphics[width=0.495\textwidth]{figures/fig_NVbox_mu_pub}}
\caption{The compression of the gel in \emph{open system} (solid lines) and in \emph{closed system} (dotted lines). 
Shadowed area limit the states with applied pressure below zero and above 5 bar.
The values $N_\cl$ are the virtual numbers of present $\cl$ ions in \emph{closed system} simulation boxes.
\label{fig: NV and CN}}

%\todoi{Replace y-axis labels by $\ncl$}
\end{figure*}

@@

@@

Such estimations are provided in the fifth column of the table, for five triplets of values $\cs^f$, $\cs^p$, and $\cs^b$.
$W_{\mathrm{sim}}$ is the energy calculated by integration of $\Pgel(\Vgel)$ dependencies~--- sum of corresponding comressions in closed and open systems, $W_{\mathrm{id}}$~--- the estimate based on the equation \refeq{eq:SEC} and $R_w$~--- is the corresponding recovery ratio.
Note that calculating the $W_{\mathrm{sim}}$ we accounted for only the work done on the gel during compression, whereas the work done by the gel itself during swelling was not taken into account.
The ratio between $W_{\mathrm{sim}}$ and $W_{\mathrm{id}}$ ranges from 3.26 to 5.43, which is comparable to that of RO.

\section{Conclusions}
We have modeled the compression of the gel in thermodynamic equilibrium with the reservoir of limited amount of water.
We have shown that in case of gel made of strong polyelectrolyte, the compression always lead to decrease of the surrounding salinity.
We have modeled the process of water desalination combining two processes: 
(1) the swelling of the gel in \emph{open} system, exchanging ions with big reservoir at constant salinity. 
(2) the compression of the gel in \emph{closed} system, when the gel affect the salinity in surrounding solution.
We estimated the energy needed to produce one liter of water and have show that the method 
%at least theoretically 
may compete with the modern technologies used in modern industry.

@@

\newpage
%\listoftodos

@@
